\RequirePackage[l2tabu]{nag}		% Warns for incorrect (obsolete) LaTeX usage

\documentclass[a4paper,12pt,openbib,oneside]{memoir} 
\usepackage{gensymb}
\usepackage{datetime}
\usepackage{ifpdf}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{mathptmx} % Times New Roman-like font
\usepackage[T1]{fontenc}

% Set margins: 1.5in left, 1in top/right/bottom
\geometry{left=1.5in, right=1in, top=1in, bottom=1in}

\ifpdf
\pdfinfo{
   /Author (Mrs. Shaima Aslam Chaity)
   /Title (Undergrad Thesis)
   /Keywords (200121)
   /CreationDate (D:\pdfdate)
}
\fi
 
\ifdraftdoc 
	\usepackage{draftwatermark}				
	\SetWatermarkScale{0.3}
	\SetWatermarkText{\bf Draft: \today}
\fi

% Declare figure/table as a subfloat.
\frenchspacing
% Set 1.5 line spacing globally
\OnehalfSpacing 
% Sets numbering division level
\setsecnumdepth{subsection} 
\maxsecnumdepth{subsubsection}

% Chapter style (taken and slightly modified from Lars Madsen Memoir Chapter Styles)
\usepackage{calc,soul,fourier}
\makeatletter 
\newlength\dlf@normtxtw 
\setlength\dlf@normtxtw{\textwidth} 
\newsavebox{\feline@chapter} 
\newcommand\feline@chapter@marker[1][4cm]{%
	\sbox\feline@chapter{% 
		\resizebox{!}{#1}{\fboxsep=1pt%
			\colorbox{gray}{\color{white}\thechapter}% 
		}}%
		\rotatebox{90}{% 
			\resizebox{%
				\heightof{\usebox{\feline@chapter}}+\depthof{\usebox{\feline@chapter}}}% 
			{!}{\scshape\so\@chapapp}}\quad%
		\raisebox{\depthof{\usebox{\feline@chapter}}}{\usebox{\feline@chapter}}%
} 
\newcommand\feline@chm[1][4cm]{%
	\sbox\feline@chapter{\feline@chapter@marker[#1]}% 
	\makebox[0pt][c]{% aka \rlap
		\makebox[1cm][r]{\usebox\feline@chapter}%
	}}
\makechapterstyle{daleifmodif}{
	\renewcommand\chapnamefont{\normalfont\Large\scshape\raggedleft\so} 
	\renewcommand\chaptitlefont{\normalfont\Large\bfseries\scshape} 
	\renewcommand\afterchapternum{\par\vskip\midchapskip} 
}
\makeatother 
\chapterstyle{daleifmodif}

% Page style: page numbers at bottom center, headers for sections
\makepagestyle{myvf} 
\makeoddfoot{myvf}{}{\thepage}{} 
\makeevenfoot{myvf}{}{\thepage}{} 
\makeheadrule{myvf}{\textwidth}{\normalrulethickness} 
\makeevenhead{myvf}{\small\textsc{\leftmark}}{}{} 
\makeoddhead{myvf}{}{}{\small\textsc{\rightmark}}
\pagestyle{myvf}

% Command to clear to next odd page
\newcommand{\clearemptydoublepage}{\newpage{\thispagestyle{empty}\cleardoublepage}}

% Creates indexes for Table of Contents, List of Figures, List of Tables, and Index
\makeindex

\usepackage{import}
\usepackage{lipsum}
\usepackage{amsfonts}
\usepackage[centertags]{amsmath}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{amsthm}
\usepackage{newlfont}
\usepackage{graphicx}
\usepackage{longtable,rotating}
\usepackage[applemac]{inputenc}
\usepackage{colortbl}
\usepackage{wasysym}
\usepackage{mathrsfs}
\usepackage{subcaption}
\usepackage{float}
\usepackage{multirow}
\usepackage{verbatim}
\usepackage{upgreek}
\usepackage{latexsym}
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage{url}
\usepackage[spanish,english]{babel}
\usepackage{color}
\usepackage[colorlinks=true,allcolors=black]{hyperref}
\usepackage{memhfixc}
\usepackage{footnote}
\usepackage{microtype}
\usepackage{rotfloat}
\usepackage{alltt}
\usepackage[version=0.96]{pgf}
\usepackage{tikz}
\usepackage{graphicx} % For figure environment
\usepackage{float}
\usepackage{seqsplit}
\usetikzlibrary{arrows,shapes,automata,backgrounds,petri,topaths, shapes.geometric, arrows.meta}

% Reduce widows and orphans
\widowpenalty=1000
\clubpenalty=1000

% Custom commands
\newcommand{\keywords}[1]{\par\noindent{\small{\bf Keywords:} #1}}
\newcommand{\parcial}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\vectorr}[1]{\mathbf{#1}}
\newcommand{\vecol}[2]{\left(\begin{array}{c} \displaystyle#1 \\ \displaystyle#2 \end{array}\right)}
\newcommand{\mados}[4]{\left(\begin{array}{cc} \displaystyle#1 &\displaystyle #2 \\ \displaystyle#3 & \displaystyle#4 \end{array}\right)}
\newcommand{\pgftextcircled}[1]{%
    \setbox0=\hbox{#1}%
    \dimen0\wd0%
    \divide\dimen0 by 2%
    \begin{tikzpicture}[baseline=(a.base)]%
        \useasboundingbox (-\the\dimen0,0pt) rectangle (\the\dimen0,1pt);
        \node[circle,draw,outer sep=0pt,inner sep=0.1ex] (a) {#1};
    \end{tikzpicture}
}
\newcommand{\range}[1]{\textnormal{range }#1}
\newcommand{\innerp}[2]{\left\langle#1,#2\right\rangle}
\newcommand{\prom}[1]{\left\langle#1\right\rangle}
\newcommand{\tra}[1]{\textnormal{tra} \: #1}
\newcommand{\sign}[1]{\textnormal{sign\,}#1}
\newcommand{\sech}[1]{\textnormal{sech} #1}
\newcommand{\diag}[1]{\textnormal{diag} #1}
\newcommand{\arcsech}[1]{\textnormal{arcsech} #1}
\newcommand{\arctanh}[1]{\textnormal{arctanh} #1}
\newcommand{\blackged}{\hfill$\blacksquare$}
\newcommand{\whiteged}{\hfill$\square$}
\newcounter{proofcount}
\renewenvironment{proof}[1][\proofname.]{\par
 \ifnum \theproofcount>0 \pushQED{\whiteged} \else \pushQED{\blackged} \fi%
 \refstepcounter{proofcount}
 \normalfont 
 \trivlist
 \item[\hskip\labelsep \itshape {\bf\em #1}]\ignorespaces
}{\addtocounter{proofcount}{-1} \popQED\endtrivlist}

% New square root definition
\let\oldsqrt\sqrt
\def\sqrt{\mathpalette\DHLhksqrt}
\def\DHLhksqrt#1#2{%
\setbox0=\hbox{$#1\oldsqrt{#2\,}$}\dimen0=\ht0
\advance\dimen0-0.2\ht0
\setbox2=\hbox{\vrule height\ht0 depth -\dimen0}%
{\box0\lower0.4pt\box2}}

% Caption styles
\newcommand{\mycaption}[2][\@empty]{%
	\captionnamefont{\scshape} 
	\changecaptionwidth
	\captionwidth{0.9\linewidth}
	\captiondelim{.\:} 
	\indentcaption{0.75cm}
	\captionstyle[\centering]{}
	\setlength{\belowcaptionskip}{10pt}
	\ifx \@empty#1 \caption{#2}\else \caption[#1]{#2}
}
\newcommand{\mysubcaption}[2][\@empty]{%
	\subcaptionsize{\small}
	\hangsubcaption
	\subcaptionlabelfont{\rmfamily}
	\sidecapstyle{\raggedright}
	\setlength{\belowcaptionskip}{10pt}
	\ifx \@empty#1 \subcaption{#2}\else \subcaption[#1]{#2}
}

% Initial letter
\usepackage{lettrine}
\newcommand{\initial}[1]{%
	\lettrine[lines=3,lhang=0.33,nindent=0em]{\color{gray}{\textsc{#1}}}{}}

% Theorem styles
\theoremstyle{plain}
\newtheorem{theo}{Theorem}[chapter]
\theoremstyle{plain}
\newtheorem{prop}{Proposition}[chapter]
\theoremstyle{plain}
\theoremstyle{definition}
\newtheorem{dfn}{Definition}[chapter]
\theoremstyle{plain}
\newtheorem{lema}{Lemma}[chapter]
\theoremstyle{plain}
\newtheorem{cor}{Corollary}[chapter]
\theoremstyle{plain}
\newtheorem{resu}{Result}[chapter]

% Hyphenation
\hyphenation{res-pec-tively}
\hyphenation{mono-ti-ca-lly}
\hyphenation{hypo-the-sis}
\hyphenation{para-me-ters}
\hyphenation{sol-va-bi-li-ty}

% new things added:

% Set no paragraph indentation and increase word spacing with justification
\setlength{\parindent}{0pt} % No indentation for paragraphs
\fontdimen2\font=0.3em % Nominal inter-word space
\fontdimen3\font=0.15em % Stretch for inter-word space
\fontdimen4\font=0.1em % Shrink for inter-word space
% \setlength{\spaceskip}{0.2em}
\setbeforesecskip{4.5ex plus 1ex minus 0.2ex} % Increases space before \section
\setbeforesubsecskip{4.25ex plus 1ex minus 0.2ex} % Increases space before \subsection

\setlength{\parskip}{12pt}

\begin{document}

\frontmatter
\pagenumbering{roman}
\input{frontmatter/title}
\input{frontmatter/Declaration.tex}
\input{frontmatter/certification}
\input{frontmatter/acknowledgment}
\input{frontmatter/abstract}

\renewcommand{\contentsname}{Table of Contents}
\maxtocdepth{subsection}
\tableofcontents*
\addtocontents{toc}{\par\nobreak \mbox{}\hfill{\bf Page}\par\nobreak}
\clearpage

\input{frontmatter/nomenclature.tex}
\clearpage

\listoftables
\addtocontents{lot}{\par\nobreak\textbf{{\scshape Table} \hfill Page}\par\nobreak}
\clearpage

\listoffigures
\addtocontents{lof}{\par\nobreak\textbf{{\scshape Figure} \hfill Page}\par\nobreak}
\clearpage

\mainmatter
\import{chapters/chapter01/}{chap01.tex}
\import{chapters/chapter02/}{chap02.tex}
\import{chapters/chapter03/}{chap03.tex}
\import{chapters/chapter04/}{chap04.tex}
\import{chapters/chapter05/}{chap05.tex}

\renewcommand\bibname{\fontsize{16pt}{1pt} \selectfont \centerline{REFERENCES} \global\def\bibname{REFERENCES}}

\begin{thebibliography}{45}

\bibitem{b1}
Malik, S.E., Kanwal, S., Javed, J., Hidayat, W., Ghaffar, T., Aamir, A.H.: Endocrine disorders in Beta-Thalassemia major patients at a Tertiary Care Hospital. \emph{Pakistan Journal of Medical Sciences} \textbf{39}(3), 726 (2023).

\bibitem{b2}
Akiki, N., Hodroj, M.H., Bou-Fakhredin, R., Matli, K., Taher, A.T.: Cardiovascular complications in $\beta$-thalassemia: getting to the heart of it. \emph{Thalassemia Reports} \textbf{13}(1), 38-50 (2023).

\bibitem{b3}
Podder, N.K., Rana, H.K., Azam, M.S., Rana, M.S., Akhtar, M.R., Rahman, M.R., Rahman, M.H., Moni, M.A.: A system biological approach to investigate the genetic profiling and comorbidities of type 2 diabetes. \emph{Gene Reports} \textbf{21}, 100830 (2020).

\bibitem{b4}
Cao, A., Galanello, R.: Beta-thalassemia. \emph{Genetics in Medicine} \textbf{12}(2), 61-76 (2010).

\bibitem{b5}
Harrison, N.A., Kopelman, M.D., David, A.S., Fleminger, S.: Endocrine diseases and metabolic disorders. In: \emph{Lishman's Organic Psychiatry: A Textbook of Neuropsychiatry}, 4th ed., pp. 617-688. Wiley-Blackwell, Chichester (2009).

\bibitem{b6}
Wuche, C.: The cardiovascular system and associated disorders. \emph{British Journal of Nursing} \textbf{31}(17), 886-892 (2022).

\bibitem{b7}
Carsote, M., Vasiliu, C., Trandafir, A.I., Albu, S.E., Dumitrascu, M.C., Popa, A., Mehedintu, C., Petca, R.C., Petca, A., Sandru, F.: New entity—thalassemic endocrine disease: major beta-thalassemia and endocrine involvement. \emph{Diagnostics} \textbf{12}(8), 1921 (2022).

\bibitem{b8}
Akiki, N., Hodroj, M.H., Bou-Fakhredin, R., Matli, K., Taher, A.T.: Cardiovascular complications in $\beta$-thalassemia: getting to the heart of it. \emph{Thalassemia Reports} \textbf{13}(1), 38-50 (2023).

\bibitem{b9}
Ali, S., Mumtaz, S., Shakir, H.A., Khan, M., Tahir, H.M., Mumtaz, S., Mughal, T.A., Hassan, A., Kazmi, S.A.R., Sadia, Irfan, M.: Current status of beta-thalassemia and its treatment strategies. \emph{Molecular Genetics \& Genomic Medicine} \textbf{9}(12), e1788 (2021).

\bibitem{b10}
Podder, N.K., Shill, P.C., Rana, H.K., Omit, S.B.S., Al Shahriar, M.M.H., Azam, M.S.: Genetic effects of COVID-19 on the development of neurodegenerative diseases. In: \emph{2021 5th International Conference on Electrical Information and Communication Technology (EICT)}, pp. 1-6. IEEE, Piscataway (2021).

\bibitem{b11}
Nain, Z., Rana, H.K., Lio, P., Islam, S.M.S., Summers, M.A., Moni, M.A.: Pathogenetic profiling of COVID-19 and SARS-like viruses. \emph{Briefings in Bioinformatics} \textbf{22}(2), 1175-1196 (2021).

\bibitem{b12}
Khare, S., Chatterjee, T., Gupta, S., Patel, A.: Molecular pathways and gene networks in fetal hemoglobin as a novel protein target for beta-thalassemia. \emph{MGM Journal of Medical Sciences} \textbf{10}(3), 485-492 (2023).

\bibitem{b13}
Sarkar, M.S., Mia, M.M., Al Amin, M., Hossain, M.S., Islam, M.Z.: Bioinformatics and network biology approach to identifying type 2 diabetes genes and pathways that influence the progression of breast cancer. \emph{Helioyon} \textbf{9}(5) (2023).

\bibitem{b14}
Nain, Z., Barman, S.K., Sheam, M.M., Syed, S.B., Samad, A., Quinn, J.M., Karim, M.M., Himel, M.K., Roy, R.K., Moni, M.A., Biswas, S.K.: Transcriptomic studies revealed pathophysiological impact of COVID-19 to predominant health conditions. \emph{Briefings in Bioinformatics} \textbf{22}(6), bbab197 (2021).

\bibitem{b15}
Datta, R., Podder, N.K., Rana, H.K., Islam, M.K.B., Moni, M.A.: Bioinformatics approach to analyze gene expression profile and comorbidities of gastric cancer. In: \emph{2020 23rd International Conference on Computer and Information Technology (ICCIT)}, pp. 1-6. IEEE, Piscataway (2020).

\bibitem{b16}
Deshpande, R., VanderSluis, B., Myers, C.L.: Comparison of profile similarity measures for genetic interaction networks. \emph{PloS One} \textbf{8}(7), e68664 (2013).

\bibitem{b17}
Kuleshov, M.V., Jones, M.R., Rouillard, A.D., Fernandez, N.F., Duan, Q., Wang, Z., Koplev, S., Jenkins, S.L., Jagodnik, K.M., Lachmann, A., McDermott, M.G.: Enrichr: a comprehensive gene set enrichment analysis web server 2016 update. \emph{Nucleic Acids Research} \textbf{44}(W1), W90-W97 (2016).

\bibitem{b18}
Podder, N.K., Shill, P.C.: Statistical and bioinformatics model to identify the influential genes and comorbidities of glioblastoma. In: \emph{International Conference on Machine Intelligence and Emerging Technologies}, pp. 763-774. Springer, Cham (2022).

\bibitem{b19}
Chen, E.Y., Tan, C.M., Kou, Y., Duan, Q., Wang, Z., Meirelles, G.V., Clark, N.R., Ma'ayan, A.: Enrichr: interactive and collaborative HTML5 gene list enrichment analysis tool. \emph{BMC Bioinformatics} \textbf{14}(1), 128 (2013).

\bibitem{b20}
Wheelock, C.E., Wheelock, A.M., Kawashima, S., Diez, D., Kanehisa, M., van Erk, M., Kleemann, R., Haeggstrom, J.Z., Goto, S.: Systems biology approaches and pathway tools for investigating cardiovascular disease. \emph{Molecular BioSystems} \textbf{5}(6), 588-602 (2009).

\bibitem{b21}
Croft, D., O'kelly, G., Wu, G., et al.: Reactome: a database of reactions, pathways and biological processes. \emph{Nucleic Acids Research} \textbf{39}(Database), D691-D697 (2011).

\bibitem{b22}
Rana, H.K., Akhtar, M.R., Islam, M.B., Ahmed, M.B., Lio, P., Quinn, J.M., Huq, F., Moni, M.A.: Genetic effects of welding fumes on the development of respiratory system diseases. \emph{Computers in Biology and Medicine} \textbf{108}, 142-149 (2019).

\bibitem{b23}
Morrison, D.A.: Networks in phylogenetic analysis: new tools for population biology. \emph{International Journal for Parasitology} \textbf{35}(5), 567-582 (2005).

\bibitem{b24}
Rana, M.S., Podder, N.K., Rana, H.K., Hasan, M.I., Azam, M.S., Rahim, M.A., Iqbal, S.H.S., Saha, S.: Identification of genomic associations between Parkinson's and neurodegenerative diseases using bioinformatics models. In: \emph{2023 International Conference on Electrical, Computer and Communication Engineering (ECCE)}, pp. 1-6. IEEE, Piscataway (2023).

\bibitem{b25}
Slavov, S.N., Haddad, S.K., Silva-Pinto, A.C., Amarilla, A.A., Alfonso, H.L., Aquino, V.H., Covas, D.T.: Molecular and phylogenetic analyses of human Parvovirus B19 isolated from Brazilian patients with sickle cell disease and $\beta$-thalassemia major and healthy blood donors. \emph{Journal of Medical Virology} \textbf{84}(10), 1652-1665 (2012).

\bibitem{b27}
Durrani, I.A., Bhatti, A., John, P.: Integrated bioinformatics analyses identifying potential biomarkers for type 2 diabetes mellitus and breast cancer: In SIK1-ness and health. \emph{PLoS One} \textbf{18}(8), e0289839 (2023).

\bibitem{b28}
Slenter, D.N., Kutmon, M., Hanspers, K., et al.: WikiPathways: a multifaceted pathway database bridging metabolomics to other omics research. \emph{Nucleic Acids Research} \textbf{46}(D1), D661-D667 (2018).

\bibitem{b29}
Barrett, T., Wilhite, S.E., Ledoux, P., et al.: NCBI GEO: archive for functional genomics data sets—update. \emph{Nucleic Acids Research} \textbf{41}(D1), D991-D995 (2012).

\bibitem{b30}
Talihati, Z., Abudurousuli, K., Hailati, S., Han, M., Nuer, M., Khan, N., ... Zhou, W.: Screening of hub genes and therapeutic drugs in cervical cancer using integrated bioinformatics analysis. \emph{Journal of Cancer} \textbf{16}(1), 92 (2025).

\bibitem{b31}
Caso, V.M., Manzo, V., Pecchillo Cimmino, T., et al.: Regulation of inflammation and oxidative stress by formyl peptide receptors in cardiovascular disease progression. \emph{Lifestyles} \textbf{11}(3), 243 (2021).

\bibitem{b32}
Ashburner, M., Ball, C.A., Blake, J.A., Botstein, D., Butler, H., Cherry, J.M., Davis, A.P., Dolinski, K., Dwight, S.S., Eppig, J.T., Harris, M.A.: Gene ontology: tool for the unification of biology. \emph{Nature Genetics} \textbf{25}(1), 25-29 (2000).

\bibitem{b33}
Podder, N.K., Shill, P.C.: Statistical and bioinformatics model to identify the influential genes and comorbidities of glioblastoma. In: \emph{International Conference on Machine Intelligence and Emerging Technologies}, pp. 763-774. Springer, Cham (2022).

\bibitem{b34}
Wishart, D.S., Feunang, Y.D., Guo, A.C., Lo, E.J., Marcu, A., Grant, J.R., Sajed, T., Johnson, D., Li, C., Sayeeda, Z., Assempour, N.: DrugBank 5.0: a major update to the DrugBank database for 2018. \emph{Nucleic Acids Research} \textbf{46}(D1), D1074-D1082 (2018).

\bibitem{b35}
Barua, J.D., Omit, S.B.S., Rana, H.K., Podder, N.K., Chowdhury, U.N., Rahman, M.H.: Bioinformatics and system biological approaches for the identification of genetic risk factors in the progression of cardiovascular disease. \emph{Cardiovascular Therapeutics} \textbf{2022}(1), 9034996 (2022).

\bibitem{b36}
Rahman, M.H., Peng, S., Hu, X., Chen, C., Uddin, S., Quinn, J.M., Moni, M.A.: Bioinformatics methodologies to identify interactions between type 2 diabetes and neurological comorbidities. \emph{IEEE Access} \textbf{7}, 183948-183970 (2019).

\bibitem{b37}
Colinge, J., Rix, U., Bennett, K.L., Superti-Furga, G.: Systems biology analysis of protein-drug interactions. \emph{PROTEOMICS-Clinical Applications} \textbf{6}(1-2), 102-116 (2012).

\bibitem{b38}
Kumar, S., Tamura, K., Nei, M.: MEGA: molecular evolutionary genetics analysis software for microcomputers. \emph{Bioinformatics} \textbf{10}(2), 189-191 (1994).

\bibitem{b39}
Rahman, M.H., Rana, H.K., Peng, S., Hu, X., Chen, C., Quinn, J.M., Moni, M.A.: Bioinformatics and machine learning methodologies to identify the effects of central nervous system disorders on glioblastoma progression. \emph{Briefings in Bioinformatics} \textbf{22}(5), bbaa365 (2021).

\bibitem{b40}
Kuleshov, M.V., Jones, M.R., Rouillard, A.D., Fernandez, N.F., Duan, Q., Wang, Z., Koplev, S., Jenkins, S.L., Jagodnik, K.M., Lachmann, A., McDermott, M.G.: Enrichr: a comprehensive gene set enrichment analysis web server 2016 update. \emph{Nucleic Acids Research} \textbf{44}(W1), W90-W97 (2016).

\bibitem{b41}
Zhou, G., Soufan, O., Ewald, J., Hancock, R.E., Basu, N., Xia, J.: NetworkAnalyst 3.0: a visual analytics platform for comprehensive gene expression profiling and meta-analysis. \emph{Nucleic Acids Research} \textbf{47}(W1), W234-W241 (2019).

\bibitem{b42}
Moni, M.A., Lio, P.: Genetic profiling and comorbidities of zika infection. \emph{The Journal of Infectious Diseases} \textbf{216}(6), 703-712 (2017).

\bibitem{b43}
Podder, N.K., Akter, F., Puza, A.K., Shill, P.C., Rana, H.K., Saha, K.K., Datta, R., Hossain, M.A.: A bioinformatics approach to identify the influences of diabetes on the progression of cancers. In: \emph{2022 International Conference on Advancement in Electrical and Electronic Engineering (ICAEEE)}, pp. 1-6. IEEE (2022).

\bibitem{b44}
Yan, S., Meng, L., Guo, X., Chen, Z., Zhang, Y., Li, Y.: Identification of ITGAX and CCR1 as potential biomarkers of atherosclerosis via gene set enrichment analysis. \emph{Journal of International Medical Research} \textbf{50}(3), 03000605211039480 (2022).

\bibitem{b45}
Liu, R., Chen, Y.: Comprehensive analysis of hub genes, biological mechanism and predicted drugs related to iron metabolism in $\beta$-thalassemia. (2023).

\end{thebibliography}

\end{document}